% Module 1: What This Is and What You Can Do With It
% Beamer Presentation — Claude Code for Academic Researchers
\documentclass[aspectratio=169, 11pt]{beamer}

% ── Theme and Colors ──────────────────────────────────────────────
\usetheme{Madrid}
\usecolortheme{default}

% Custom colors
\definecolor{anthropic}{RGB}{50, 50, 80}
\definecolor{accent}{RGB}{180, 90, 40}
\definecolor{lightgray}{RGB}{245, 245, 245}

\setbeamercolor{structure}{fg=anthropic}
\setbeamercolor{frametitle}{bg=anthropic, fg=white}
\setbeamercolor{title}{bg=anthropic, fg=white}
\setbeamercolor{block title}{bg=anthropic, fg=white}
\setbeamercolor{block body}{bg=lightgray}
\setbeamercolor{alerted text}{fg=accent}

\setbeamerfont{title}{size=\Large, series=\bfseries}
\setbeamerfont{frametitle}{size=\large, series=\bfseries}

% ── Packages ──────────────────────────────────────────────────────
\usepackage{graphicx}
\usepackage{booktabs}
\usepackage{listings}
\usepackage{fontenc}
\usepackage{hyperref}
\usepackage{tikz}

\lstset{
  basicstyle=\ttfamily\small,
  backgroundcolor=\color{lightgray},
  frame=single,
  framerule=0pt,
  breaklines=true,
  columns=fullflexible,
}

% ── Metadata ──────────────────────────────────────────────────────
\title{Claude Code for Academic Researchers}
\subtitle{A Methods Workshop in AI-Assisted Research}
\author{Brady Allardice}
\date{}
\institute{Universitat Pompeu Fabra \& Institut Barcelona d'Estudis Internacionals (IBEI)}

% ══════════════════════════════════════════════════════════════════
\begin{document}

% ── Course Title ──────────────────────────────────────────────────
\begin{frame}[plain]
  \titlepage
\end{frame}

% ── About the Instructor ─────────────────────────────────────────
\begin{frame}{About Me}
  \begin{columns}[T]
    \begin{column}{0.52\textwidth}
      \textbf{Brady Allardice}\\
      PhD Candidate, Political \& Social Sciences\\
      UPF / IBEI, Barcelona

      \vspace{0.5em}
      \textbf{Research:} Political economy of technological change
      \begin{itemize}
        \item Automation, AI, and electoral behavior
        \item LLM-based text classification pipelines
        \item Causal inference with panel data
      \end{itemize}

      \vspace{0.5em}
      \textbf{Background:}
      \begin{itemize}
        \item M.Sc.\ Economics (Barcelona School of Economics)
        \item Former data scientist (Paysafe Group)
        \item Former research associate (UC Riverside)
      \end{itemize}
    \end{column}
    \begin{column}{0.44\textwidth}
      \textbf{Why this course:}
      \begin{itemize}
        \item Daily user of Claude Code for academic research
        \item Use it across the full pipeline: data cleaning, analysis, robustness checks, code auditing, replication
        \item These tools can dramatically expand what a researcher can accomplish---if used with the right workflows
      \end{itemize}

      \vspace{0.5em}
      {\small brady.allardice@upf.edu\\github.com/bradyall}
    \end{column}
  \end{columns}
\end{frame}

% ── Course Overview ───────────────────────────────────────────────
\begin{frame}{Today's Plan}
  \small
  \begin{tabular}{@{}r l l l@{}}
    \toprule
    \textbf{\#} & \textbf{Module} & \textbf{Time} & \textbf{Format} \\
    \midrule
    1 & What This Is and What You Can Do With It & 50 min & Demo + Exercise \\
    2 & How Claude Code Works & 75 min & Lecture + Exercise \\
    3 & Automating Research Tasks & 60 min & Exercise-Heavy \\
    4 & Robustness and Replication & 60 min & Exercise-Heavy \\
    \midrule
      & \textit{Break} & \textit{15 min} & \\
    \midrule
    5 & Code Auditing and Review & 60 min & Exercise + Discussion \\
    6 & Version Control with Git & 45 min & Exercise-Heavy \\
    7 & Developing Expert Judgment & 40 min & Group Discussion \\
    8 & The Anthropic API (Bonus) & 35 min & Demo + Exercise \\
    9 & Capstone Project & 90 min & Independent Work \\
    \bottomrule
  \end{tabular}

  \vspace{0.5em}
  {\footnotesize ${\sim}$8.5 hours \quad$\vert$\quad 60--70\% hands-on \quad$\vert$\quad Pair programming throughout}
\end{frame}

\begin{frame}{Core Framework}
  \begin{alertblock}{The Central Idea}
    You are the principal investigator. Claude Code is the research assistant.\\
    You direct the work, make analytical decisions, and verify every result.\\
    Claude Code handles the mechanical execution.
  \end{alertblock}

  \vspace{0.5em}
  The question is never ``can Claude Code do this''---it almost always can.

  \vspace{0.3em}
  The question is: \textbf{what is the most productive and reliable way to get this done?}

  \vspace{0.8em}
  \textbf{By the end of today, you will be able to:}
  \begin{itemize}
    \item Automate data cleaning, robustness checks, code auditing, and documentation
    \item Choose the right execution mode for each task
    \item Verify output against known properties of your data
    \item Recognize when Claude Code is failing and intervene effectively
  \end{itemize}
\end{frame}

% ══════════════════════════════════════════════════════════════════
%  MODULE 1 BEGINS
% ══════════════════════════════════════════════════════════════════

\begin{frame}[plain]
  \centering
  \vfill
  {\Large\textbf{Module 1}}\\[0.5em]
  {\large What This Is and What You Can Do With It}\\[0.3em]
  {\small 50 minutes \quad$\vert$\quad Demo + Exercise}
  \vfill
\end{frame}

% ══════════════════════════════════════════════════════════════════
%  OPENING DEMO  [0:00–2:00]
% ══════════════════════════════════════════════════════════════════

\begin{frame}{Live Demo: The Data}
  \textbf{Dataset 1: County Presidential Election Returns}
  \begin{itemize}
    \item Source: MIT Election Data + Science Lab
    \item County-level vote totals by candidate and party, 2000--2024
    \item ${\sim}$94{,}000 rows --- one row per county $\times$ year $\times$ candidate
    \item Variables: FIPS codes, candidate names, party, vote counts, vote mode
  \end{itemize}

  \vspace{0.5em}
  \textbf{Dataset 2: IPUMS American Community Survey Microdata}
  \begin{itemize}
    \item Source: IPUMS USA (University of Minnesota)
    \item Person-level survey records with demographic, economic, and geographic variables
    \item ${\sim}$15 million rows per year --- one row per person
    \item Variables: age, education, employment, income, race, poverty status, county FIPS
  \end{itemize}
\end{frame}

\begin{frame}{Live Demo: The Task}
  \textbf{Goal:} Build a county-year panel joining election results with demographics.

  \vspace{0.5em}
  This requires:
  \begin{enumerate}
    \item Reshape election data to one row per county-year with vote shares
    \item Aggregate 15M person-level records to county-level means using survey weights
    \item Harmonize FIPS codes across datasets (different formats)
    \item Merge on county $\times$ year and validate the join
  \end{enumerate}

  \vspace{0.5em}
  In Stata or R, this takes 30--45 minutes of careful coding.

  {\footnotesize\textit{[Switch to terminal]}}
\end{frame}

% ══════════════════════════════════════════════════════════════════
%  WHAT JUST HAPPENED  [2:00–10:00]
% ══════════════════════════════════════════════════════════════════

\begin{frame}{What Just Happened}
  \begin{columns}[T]
    \begin{column}{0.48\textwidth}
      \textbf{It wrote the code:}
      \begin{itemize}
        \item Read both files
        \item Identified the join key
        \item Handled cleaning steps
        \item Wrote a complete script
      \end{itemize}
    \end{column}
    \begin{column}{0.48\textwidth}
      \textbf{It also:}
      \begin{itemize}
        \item Understood the data structure\\
          {\small (person-level vs.\ county-level,\\survey weights, FIPS formats)}
        \item Ran the code and diagnosed errors
        \item Provided summaries of the output\\
          {\small (row counts, variable descriptions,\\merge diagnostics)}
      \end{itemize}
    \end{column}
  \end{columns}

  \vspace{0.5em}
  \alert{But it also made a decision about our data without asking.}

  \vspace{0.5em}
  Both of these things are true:
  \begin{enumerate}
    \item It took 90 seconds instead of 45 minutes
    \item It made analytical choices autonomously
  \end{enumerate}
\end{frame}

\begin{frame}{The RA Analogy}
  \begin{block}{How to Think About Claude Code}
    Claude Code is like a research assistant who is:
    \begin{itemize}
      \item Extremely fast
      \item Has read a lot
      \item Will \textbf{never} say ``I don't know how to do that''
    \end{itemize}
    \vspace{0.3em}
    Instead, it will just do it wrong and hand it in.
  \end{block}

  \vspace{0.5em}
  You have all managed RAs. You know how valuable---and how dangerous---that profile is.

  \vspace{0.5em}
  \textbf{This course teaches you to capture the value while managing the risk.}
\end{frame}

% ══════════════════════════════════════════════════════════════════
%  THE CODING PARADIGM SPECTRUM  [10:00–16:00]
% ══════════════════════════════════════════════════════════════════

\begin{frame}{The Coding Paradigm Spectrum}
  \begin{tikzpicture}[
    node distance=0.3cm,
    every node/.style={draw, rounded corners, minimum width=10cm, minimum height=0.9cm, align=center, font=\small},
    arrow/.style={->, thick, >=stealth}
  ]
    \node (manual) at (0,3.6) {\textbf{Manual Coding} --- You write every line (Stata/R today)};
    \node (copilot) at (0,2.4) {\textbf{Autocomplete / Copilot} --- AI suggests; you accept or reject};
    \node (chat) at (0,1.2) {\textbf{Web Chat} --- You describe; AI gives code to copy-paste};
    \node[fill=anthropic!15] (agentic) at (0,0) {\textbf{Agentic Coding} --- AI reads, writes, and runs code in your project};
  \end{tikzpicture}

  \vspace{0.5em}
  \centering
  \small $\uparrow$ More autonomy, more power, more risk $\uparrow$
\end{frame}

\begin{frame}{Why Agentic Coding Is Different}
  \begin{alertblock}{The Key Insight}
    Agentic coding is the first paradigm where the tool \textbf{modifies your project directly}.
  \end{alertblock}

  \vspace{0.5em}
  That is what makes it so productive---and why learning to use it well matters.

  \vspace{0.5em}
  These principles apply to Claude Code, Cursor, Windsurf, or whatever comes next.
\end{frame}

% ══════════════════════════════════════════════════════════════════
%  WHAT IT CAN AND CANNOT DO  [16:00–19:00]
% ══════════════════════════════════════════════════════════════════

\begin{frame}{What Claude Code Can Do}
  \begin{columns}[T]
    \begin{column}{0.48\textwidth}
      \textbf{Delegate freely:}
      \begin{itemize}
        \item Write code rapidly
        \item Clean data
        \item Generate robustness checks
        \item Create visualizations
      \end{itemize}
    \end{column}
    \begin{column}{0.48\textwidth}
      \textbf{Delegate freely (cont.):}
      \begin{itemize}
        \item Audit existing code
        \item Write documentation
        \item Build replication packages
        \item Process text at scale
      \end{itemize}
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{What Requires Human Judgment}
  \begin{block}{The Referee Test}
    If a referee questioned this choice, who would need to answer---you or the RA?
  \end{block}

  \vspace{0.5em}
  \textbf{Do not delegate:}
  \begin{itemize}
    \item Choosing specifications
    \item Interpreting results
    \item Identification strategies
    \item Sample selection decisions
  \end{itemize}
\end{frame}

% ══════════════════════════════════════════════════════════════════
%  WHY VERIFICATION MATTERS  [19:00–23:00]
% ══════════════════════════════════════════════════════════════════

\begin{frame}{Why Verification Matters}
  {\large What happens when you skip verification:}

  \vspace{0.8em}
  \begin{enumerate}
    \item \textbf{Silent observation loss:} A merge silently drops 300 observations.\\
      The code looks clean. The error is invisible without checking row counts.

      \vspace{0.5em}
    \item \textbf{Invalid transformation:} A log transformation applied to a variable containing zeros.\\
      No error raised, wrong results.

      \vspace{0.5em}
    \item \textbf{Incorrect panel operations:} A lagged variable computed without sorting by time within panels.
  \end{enumerate}

  \vspace{0.5em}
  \alert{These are the kinds of errors that end up in published work.}
\end{frame}

% ══════════════════════════════════════════════════════════════════
%  THE CONTROL PRINCIPLE  [23:00–25:00]
% ══════════════════════════════════════════════════════════════════

\begin{frame}{The Control Principle}
  \begin{alertblock}{You Are the PI. Claude Code Is the RA.}
    The question is not ``can Claude Code do this'' but\\
    ``what is the best way to get this done reliably.''
  \end{alertblock}

  \vspace{1em}
  \centering
  \textbf{If you cannot verify a result, you should not delegate the task.}

  \vspace{1em}
  That is the standard we hold ourselves to today.
\end{frame}

% ══════════════════════════════════════════════════════════════════
%  WHY PYTHON  [25:00–27:00]
% ══════════════════════════════════════════════════════════════════

\begin{frame}{Why Python}
  \begin{itemize}
    \item Claude Code is substantially better at Python than Stata or R
    \item The principles---task decomposition, verification, control---transfer to any language
    \item Python is increasingly where the tools and ecosystem are for computational research
  \end{itemize}
\end{frame}

% ══════════════════════════════════════════════════════════════════
%  SETUP  [27:00–37:00]
% ══════════════════════════════════════════════════════════════════

\begin{frame}{Setup Check}
  \begin{enumerate}
    \item Open your terminal
    \item Verify Node.js: \lstinline{node --version}
    \item Verify Claude Code: \lstinline{claude --version}
    \item Authenticate: \lstinline{claude}
  \end{enumerate}

  \vspace{1em}
  \textbf{If you completed pre-work:} help your neighbor.\\
  \textbf{If you're stuck:} pair with someone who is set up---we'll troubleshoot at break.
\end{frame}

% ══════════════════════════════════════════════════════════════════
%  EXERCISE  [37:00–47:00]
% ══════════════════════════════════════════════════════════════════

\begin{frame}{Exercise: Your First Commands}
  \textbf{Task 1} (low-stakes):\\
  Here is a Python script that cleans a dataset.\\
  Ask Claude Code to explain what it does, line by line.

  \vspace{0.8em}
  \textbf{Task 2} (introduces verification):\\
  Ask Claude Code to add an assertion that the row count after cleaning matches the expected count.

  \vspace{0.8em}
  \textbf{Stretch:}\\
  Ask Claude Code to handle a new edge case---negative values in a column that should be positive.
\end{frame}

\begin{frame}{Exercise Debrief}
  \begin{itemize}
    \item What did it get right?
    \item Anything wrong or confusing?
    \item That is normal and expected
  \end{itemize}
\end{frame}

% ══════════════════════════════════════════════════════════════════
%  PREVIEW  [47:00–50:00]
% ══════════════════════════════════════════════════════════════════

\begin{frame}{What's Ahead Today}
  By the end of the day, you will have:
  \begin{itemize}
    \item Built a documented data pipeline
    \item Generated robustness checks
    \item Audited code for errors
    \item Learned version control for agent-assisted work
  \end{itemize}

  \vspace{0.5em}
  You will know which parts Claude Code should handle and which require your judgment.

  \vspace{0.5em}
  \textbf{And you will be able to take these workflows back to your own research.}
\end{frame}

\end{document}
